---
title: "p8105_hw3_ajt2206"
author: "Andy Turner"
date: "2023-10-06"
output: github_document
---

## Problem 1:

**Loading data from library**
```{r}
library(tidyverse)
library(p8105.datasets)
data("instacart")
```
**Description of dataset and illustrative example**
The Instacart dataset is an anonymized dataset from 2017 which details over 3 milion online grocery orders from more than 200,000 Instacart Users. Despite its large scale, the dataset still represents a random sample of the overall Instacart user base. There are `r nrow(instacart)` observations, each representing an order, and `r ncol(instacart)` variables in the dataset. The data is currently in horizontal form. Some of the key variables to note are: 

* `order_id`: order identifier
* `product_id`: product identifier
* `user_id`: customer identifier
* `product_name`: name of the product ordered
* `order_number`: the order sequence number for this user (1=first, n=nth)
* `days_since_prior_order`: days since the last order, capped 30, NA if `order_number`=1

For example, Observation 58 describes the 26th item added to the Instacart order (`add_to_cart_order`= 26) by user 56463 (`user_id`=56463). This is the 41st order by the user (`order_number=41`). Their prior order was 14 days/2 weeks prior (`days_since_prior_order`=14. The specific item ordered (`product_id`=47333) was Queso Fresco (`product_name`= "Queso Fresco") found in the specialty cheese aisle (`aisle`= "specialty cheese") within the dairy eggs department (`department`= "dairy eggs"). The user has ordered this item in the past (`reordered=1`)

How many aisles are there, and which aisles are the most items ordered from?

There are 134 aisles. 

Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.

Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. Include the number of times each item is ordered in your table.
Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table).

## Problem 2

Loading the BRFSS data from our p8105 library. 

```{r}
data("brfss_smart2010")

brfss=
  brfss_smart2010|> 
  janitor::clean_names() |> 
  rename(state = locationabbr) |> 
  filter(topic== "Overall Health") |> 
  mutate(response= factor(response, levels = c("Poor", "Fair", "Good", "Very good", "Excellent"))) |> 
  arrange(response)
  
```

**Data Import Cleaning Steps**

I imported the data directly into R from the p8105.datasets library which we already loaded for Question 1. Afterwards, I used the `janitor::clean_names()` to make all the variable names lower cased and replace any spaces with `_`. I renamed the locationabbr variable to state. Additionally, I filtered the dataset to only include observations where the topic was "Overall Health." Lastly, I converted the response data from character to factor using the `factor()` function in R. I also arranged the dataset using `arrange` by the response variable with the recently ordered factor. 

Using this dataset, do or answer the following (commenting on the results of each):

```{r}

```
In 2002, which states were observed at 7 or more locations? What about in 2010?

```{r}
library(ggplot2)
  
excellent_df=
  brfss |> 
  filter(response== "Excellent") |> 
  group_by(year, state) |> 
  mutate(mean_datavalue= mean(data_value)) |> 
  select (year, state, mean_datavalue) |> 
  distinct(year, state, mean_datavalue, .keep_all = TRUE)

ggplot(excellent_df, aes(x= year, y= mean_datavalue, group=state, color=state))+
  geom_line(alpha=0.5)+
  labs(
    title = "Spaghetti Plot of Mean by State",
    x = "Year",
    y = "Mean"
  ) +
  theme_minimal()
```
**Excellent Dataset and Spaghetti Plot**

To construct the excellent dataset, I filtered brfss to only include values of "Excellent" for the response variable. I grouped the data by both year and state which allowed me to use the mutate function to create a new variable "mean_datavalue" which was the average of the data_value for each state separated by year. Afterwards, I used the `select()` function to select only the variables year, state, and mean_datavalue. Last, I used the `distinct()` function to only keep unique rows. 

To make a "spaghetti" plot, I used `ggplot` specifying year on my x-axis, mean_datavalue on the y-axis, grouped by state and with each state as a different color. Moreover, I used `geom_line` to specify that I wanted a line graph. Other than that I used a few more functions to improve the aesthetics of the graph such as adding a title via the `labs()` function. 

```{r}
library(patchwork)

dv_2006 =
  brfss |> 
  filter(year== "2006", state== "NY") |> 
  ggplot(aes(x=response, y=data_value, color=locationdesc))+
  geom_point(alpha=1.0)

dv_2010 =
  brfss |>  
  filter(year== "2010", state== "NY") |> 
  ggplot(aes(x=response, y=data_value, color=locationdesc))+
  geom_point(alpha=1.0)
  
dv_2006+dv_2010
```
**Two-panel plot of data_value**
To create a two-panel plot, I first loaded the patchwork library using `library()`. Next, I worked to create the plot for 2006. I filtered the brfss dataframe to only include year as 2006 and state as NY. Next, I used `ggplot` specifying that I want response as my x-value, data-value as my y value, and to color the points by locationdesc. Lastly, I used `geom_point`to tell R to make a scatterplot. I did the same steps for the 2010 plot except, I filtered for 2010. Last, to show the two plots side by side, I just wrote dv_2006+dv_2010`` (graph_1 + graph_2) which using patchwork puts the two graphs next to each other!
Make a two-panel plot showing, for the years 2006, and 2010, distribution of data_value for responses (“Poor” to “Excellent”) among locations in NY State.