---
title: "p8105_hw3_ajt2206"
author: "Andy Turner"
date: "2023-10-06"
output: github_document
---

## Problem 1:

**Loading data from library**
```{r}
library(tidyverse)
library(p8105.datasets)
data("instacart")
```
**Description of dataset and illustrative example**
The Instacart dataset is an anonymized dataset from 2017 which details over 3 milion online grocery orders from more than 200,000 Instacart Users. Despite its large scale, the dataset still represents a random sample of the overall Instacart user base. There are `r nrow(instacart)` observations, each representing an order, and `r ncol(instacart)` variables in the dataset. The data is currently in horizontal form. Some of the key variables to note are: 

* `order_id`: order identifier
* `product_id`: product identifier
* `user_id`: customer identifier
* `product_name`: name of the product ordered
* `order_number`: the order sequence number for this user (1=first, n=nth)
* `days_since_prior_order`: days since the last order, capped 30, NA if `order_number`=1

For example, Observation 58 describes the 26th item added to the Instacart order (`add_to_cart_order`= 26) by user 56463 (`user_id`=56463). This is the 41st order by the user (`order_number=41`). Their prior order was 14 days/2 weeks prior (`days_since_prior_order`=14. The specific item ordered (`product_id`=47333) was Queso Fresco (`product_name`= "Queso Fresco") found in the specialty cheese aisle (`aisle`= "specialty cheese") within the dairy eggs department (`department`= "dairy eggs"). The user has ordered this item in the past (`reordered=1`)

How many aisles are there, and which aisles are the most items ordered from?

There are 134 aisles. 

Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.

Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. Include the number of times each item is ordered in your table.
Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table).

## Problem 2

Loading the BRFSS data from our p8105 library. 

```{r}
data("brfss_smart2010")

brfss=
  brfss_smart2010|> 
  janitor::clean_names() |> 
  rename(state = locationabbr) |> 
  filter(topic== "Overall Health")
```

**Data Import Cleaning Steps**

I imported the data directly into R from the p8105.datasets library which we already loaded for Question 1. Afterwards, I used the `janitor::clean_names()` to make all the variable names lower cased and replace any spaces with `_`. I renamed the locationabbr variable to state. Additionally, I filtered the dataset to only include observations where the topic was "Overall Health." Lastly, [ADD SENTENCE WHEN YOU SUCCESSFULLY DO THE VECTOR THING]

 --> organize responses as a factor taking levels ordered from “Poor” to “Excellent”


Using this dataset, do or answer the following (commenting on the results of each):

```{r}
brfss |> 
  group_by(locationabbr) |> 
  unique(locationdesc)

library(ggplot2)
  
excellent_df=
  brfss |> 
  filter(response== "Excellent") |> 
  group_by(year, state) |> 
  mutate(mean_datavalue= mean(data_value)) |> 
  select (year, state, mean_datavalue) |> 
  distinct(year, state, mean_datavalue, .keep_all = TRUE)

```


```{r}
ggplot(excellent_df, aes(x= year, y= mean_datavalue, group=state, color=state))+
  geom_line(alpha=0.5)+
  labs(
    title = "Spaghetti Plot of Mean by State",
    x = "Year",
    y = "Mean"
  ) +
  theme_minimal()
```

In 2002, which states were observed at 7 or more locations? What about in 2010?

Construct a dataset that is limited to Excellent responses, and contains, year, state, and a variable that averages the data_value across locations within a state. Make a “spaghetti” plot of this average value over time within a state (that is, make a plot showing a line for each state across years – the geom_line geometry and group aesthetic will help). COMPLETED

Make a two-panel plot showing, for the years 2006, and 2010, distribution of data_value for responses (“Poor” to “Excellent”) among locations in NY State.